---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# toweR

<!-- badges: start -->
[![R-CMD-check](https://github.com/ltdroy/toweR/workflows/R-CMD-check/badge.svg)](https://github.com/ltdroy/toweR/actions)
[![test-coverage](https://github.com/ltdroy/toweR/workflows/test-coverage/badge.svg)](https://github.com/ltdroy/toweR/actions)
[![Project Status: WIP – Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
[![license](https://img.shields.io/github/license/mashape/apistatus.svg)](https://choosealicense.com/licenses/mit/)
[![codecov](https://codecov.io/gh/ltdroy/toweR/branch/master/graph/badge.svg)](https://codecov.io/gh/ltdroy/toweR)
<!-- badges: end -->

The goal of toweR is to make the job of loading and vertically combining multiple rectangular data-files more efficient, easier and cleaner. 

Most importantly, it provides specialized tools for dealing with *inconsistencies in the structure of the data files* (e.g. different column names).

## Quick-Start: Tower Workflow

The basic `toweR` workflow is:

1. Build a 'file scaffold' (a special df/tibble) containing the metadata about your files
2. Use this metadata to read each file into your scaffold
3. Harmonise these files prior to combining them
4. Combine them vertically (a.k.a row-bind/vertical-merge) into a single dataframe
5. Check the diagnostics for this vertical merge
6. (As needed) repeat steps (3-5) to refine the harmonisation

### Build the file scaffold

* `build_scaffolding_from_dir()`

### Add metadata about your files

* `metadata_col_from_regex()`
* `metadata_col_from_lookup()`
* `add_linestarts()`
* `modify_linestarts()`
* `add_sheet_selections()`
* `modify_sheet_selections()`

### Check the file coverage using the metadata

* `check_metadata_coverage()`

### Load the raw data into your scaffold

* `load_materials_simply()`

### Harmonise the files prior to vertical merge

* `harmonise_by_explicit()`

### Vertically combine the the datasets (with optional diagnostics)

* `vertical_merge()`

### Extract detailed diagnostics from the combined dataframe

* `get_tower_diagnostics()`

## Installation

You can install the development version of toweR like so:

``` r
devtools::install_github("ltdroy/toweR")
```

## Background

A *vertical merge* (a.k.a a row-bind) is different to a *join*, in a *join* we are usually combining two datasets with different types (and/or) levels of measurement (e.g. the demographic characteristics of UK households, and the labor-market characteristics of UK regions), horizontally on a shared key (e.g. UK regional location).

In a vertical merge (also called vertical bind or row-bind), we are taking two or more data-sets with (in principle) the same measurement structure, that cover *different cases* (and/or the same cases measured at different times) and we are then stacking them on top of each other (hence the 'tower' metaphor). 

A trivial (and familiar to most users) example of vertical merger in `R` is a call to `rbind()`.

```{r}

df1 <- tibble::tribble(
  ~ x, ~y, 
  1, "a",
  2, "b",
  3, "c",
  4, "d",
  5, "e"
)

df2 <- tibble::tribble(
  ~ x, ~y, 
  6, "f",
  7, "g",
  8, "h",
  9, "i",
  10, "j"
)

rbind(df1, df2)

```

This simple example doesn't require any additional tooling.

Further more, `dplyr::bind_rows()` provides a useful extension of `rbind()` that accepts a list of data-frames and binds them vertically by column name. So the case of `n > 2` doesn't *necessarily* pose a problem for vertical merges (but see below).

```{r}

df1 <- tibble::tribble(
  ~ x, ~y, 
  1, "a",
  2, "b",
  3, "c",
  4, "d",
  5, "e"
)

df2 <- tibble::tribble(
  ~ x, ~y, 
  6, "f",
  7, "g",
  8, "h",
  9, "i",
  10, "j"
)

dplyr::bind_rows(list(df1, df2, df2))


```

The `toweR` package comes into its own when it comes to dealing with multiple datasets (i.e. any more than 3) that may have *inconsistent structure*:

1. Different column names, or missing/inconsistent columns
2. Different column types (e.g. date value in one dataset, character in another)
3. Unexpected missing-data patterns in some dataframes
4. Data is stored in different parts of text/excel-files

These inconsistencies can create significant problems when seeking to combine the data for scientific analyses:

**Example 1:** Inconsistently named, or missing, columns can create duplicated columns after a vertical merge, or lead to columns being silently (and potentially unexpectedly) padded with missing values for sets of cases that didn't include that column. 

```{r}

df1 <- tibble::tribble(
  ~ x, ~y, 
  1, "a",
  2, "b",
  3, "c",
  4, "d",
  5, "e"
)

df2 <- tibble::tribble(
  ~ x, ~Y, 
  6, "f",
  7, "g",
  8, "h",
  9, "i",
  10, "j"
)

dplyr::bind_rows(list(df1, df2))

```

**Example 2:** Inconsistent column types (i.e. numeric, character, etc.) can lead to unexpected coercion of columns into undesirable formats.

```{r}

df1 <- tibble::tribble(
  ~ x, ~y, 
  "1", "a",
  "2", "b",
  "3", "c",
  "4", "d",
  "5", "e"
)

df2 <- tibble::tribble(
  ~ x, ~y, 
  6, "f",
  7, "g",
  8, "h",
  9, "i",
  10, "j"
)

combined_df <- rbind(df1, df2)

combined_df

mean(combined_df$x)


```

**Example 3:** Missing data patterns can mask the absence of information in certain files. For example, a column named `x` might be present in all files in a set, but the user might not be aware that in files 4 and 19, the column `x` is filled with only missing data. 

In such an example, once the data is merged together, the absence of this information in certain files may be mistaken for a general 'low-level' presence of missing observations in the variable in question (rather than complete absence of the measure for some datasets). 

This, in turn may lead to inappropriate analytic decisions (like treating `x` as missing at random, as opposed to missing from particular datasets) and/or produce unexpected results (like attempting to calculate the mean of `x` for groups in which it was not measured at all).

```{r}

df1 <- tibble::tribble(
  ~ x, ~y, ~z,
  1, "a", 1,
  2, "b", 1,
  3, "c", 1,
  4, "d", 1,
  5, "e", 1
)

df2 <- tibble::tribble(
  ~ x, ~y, 
  6, "f",
  7, "g",
  8, "h",
  9, "i",
  10, "j"
)

df3 <- tibble::tribble(
  ~ x, ~Y, 
  NA, "f",
  NA, "g",
  NA, "h",
  NA, "i",
  NA, "j"
)

dplyr::bind_rows(list(df1, df2))

```


The `toweR` package provides a set of specialized tools for managing these issues (and others, such as inconsistent location of the data in the source files, see below).

Without specialized tools to deal with complex vertical merges, analysts and researchers may be forced to manually check for these issues and correct them. In the worst case scenario they may try to do this by hand, via 'copy and paste' between files and spreadsheets, with significant risk of manual error. 

In real-world scenarios, datasets may contain hundreds of variables and there may be tens or hundreds of datasets to combine. Manual combination quickly becomes untenable at these scales and comes with an increasing likelihood of human error. 

## Vertical join scenarios

1. A trial has been conducted in a large number of labs/hospitals/establishments across the country, and the results for each participant have been recorded in an individual spreadsheet for each establishment and then emailed to a central analysis office. Although similar, the structure of these spreadsheets has not been strongly enforced. 
2. A long-term longitudinal survey has been repeated over multiple waves, with the administrators of the survey (and therefore the format of the collated data) changing subtly over time. 
3. A researcher is scraping quarterly published data from the last ten years from the website of public body, they want to combine the 40 published datasets, but the format of this data has changed over time.
4. *Any R user that has data spread across multiple datasets or data-frames that they want to combine together quickly, easily and safely*








